---
title: "Applied Longitudinal Data Analysis in R"
subtitle: "HPD 685 — Worked Examples: MLM Growth Models & Measurement Invariance"
author: ""
date: last-modified
date-format: "MMMM DD, YYYY"
lang: en
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Overview

This document provides two fully-worked applied examples drawn from course datasets:

1.  **Part I — MLM Growth Model with Moderation** (`antisocial.csv`): A multilevel linear growth model of children's antisocial behavior across ages 8–14, testing whether sex and home cognitive environment moderate the trajectory of change.

2.  **Part II — Longitudinal Measurement Invariance** (`selfconcept.dat`): A step-by-step configural → weak → strong → partial scalar invariance sequence for a four-indicator self-concept scale measured at ages 8, 10, and 12.

------------------------------------------------------------------------

# Part I: MLM — Antisocial Behavior Growth Model with Moderation

## Background

**Dataset**: `antisocial.csv` (N = 405 children from the NLSY, National Longitudinal Surveys of Youth)

**Outcome**: Antisocial behavior (composite score, higher = more antisocial)

**Time structure**: Four assessments recorded at ages that vary by individual (ages 6–14), converted to a centered age variable

**Level-2 predictors**: `male` (0 = girl, 1 = boy); `homecog` (home cognitive support, standardized)

**Research Questions**:

-   Q1: Is there a significant mean growth trend in antisocial behavior across childhood?
-   Q2: Does the trajectory differ by sex (level and rate of change)?
-   Q3: Does the home cognitive environment predict both initial level and rate of change?

## Setup and Data Preparation

```{r}
#| label: setup
library(lme4)          # mixed-effects models
library(lmerTest)      # Satterthwaite df and p-values for lmer
library(nlme)          # heteroscedastic / multi-group lme models
library(interactions)  # sim_slopes(), interact_plot()
library(ggplot2)       # visualization
library(psych)         # descriptive statistics
library(performance)   # icc()
library(dplyr)
```

```{r}
#| label: load-data
antisocial <- read.csv("data/antisocial.csv", header = TRUE)

# Rename time-stamped columns for reshape()
colnames(antisocial) <- gsub("^age(\\d+)$",  "age_\\1",  colnames(antisocial))
colnames(antisocial) <- gsub("^anti(\\d+)$", "anti_\\1", colnames(antisocial))

# Wide → long format
antisocial_long <- reshape(
  antisocial,
  varying   = c("age_86","age_88","age_90","age_92",
                "anti_86","anti_88","anti_90","anti_92"),
  # Continuous measure of child's antisocial behavior reported by the mother in 1986, 1988, 1990, 1992
  direction = "long",
  idvar     = "id",
  sep       = "_"
)
antisocial_long <- antisocial_long[order(antisocial_long$id), ]

# Center age: subtract 6 so age=0 corresponds to ~age 6 at first wave
antisocial_long$age <- antisocial_long$age - 6

# Remove rows with missing age or outcome
antisocial_long <- antisocial_long[
  !is.na(antisocial_long$age) & !is.na(antisocial_long$anti), ]

head(antisocial_long, 12)
```

### Descriptive Statistics

```{r}
#| label: descriptives
# Sample size
cat("Total persons:", length(unique(antisocial_long$id)), "\n")
cat("Total observations:", nrow(antisocial_long), "\n\n")

# Mean antisocial behavior by age
means <- aggregate(anti ~ age, data = antisocial_long, mean)
print(means)

# Observed mean trajectory
ggplot(means, aes(age, anti)) +
  geom_point(size = 3) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue",
              linetype = "dashed") +
  labs(
    x     = "Centered Age (0 = age 6)",
    y     = "Mean Antisocial Behavior",
    title = "Observed Mean Trajectory of Antisocial Behavior"
  ) +
  theme_minimal()
```

## Step 1 — Null Model: Quantifying Clustering (ICC)

Before fitting growth trajectories, we first determine how much between-person variance exists and whether MLM is warranted. The null model contains no predictors—it simply partitions total variance into between-person ($\tau_{00}$) and within-person ($\sigma^2$) components.

$$y_{it} = \mu_0 + b_{0i} + \varepsilon_{it}$$

```{r}
#| label: null-model
m0 <- lmer(anti ~ 1 + (1 | id), data = antisocial_long)
summary(m0)
```

```{r}
#| label: icc
icc(m0)
```

**Interpretation**: The ICC indicates what proportion of total variance in antisocial behavior is attributable to stable between-person differences. An ICC \> 0.10 provides strong justification for MLM; ordinary regression would produce biased standard errors by ignoring this clustering.

------------------------------------------------------------------------

## Step 2 — Unconditional Growth Model

We add a fixed and random effect of age to model individual linear growth trajectories. Each child now has their own intercept (level at age 6) and slope (rate of annual change).

$$y_{it} = \mu_0 + \mu_1 \cdot \text{age}_{it} + b_{0i} + b_{1i}\cdot\text{age}_{it} + \varepsilon_{it}$$

```{r}
#| label: growth-model
m1 <- lmer(anti ~ age + (1 + age | id), data = antisocial_long)
summary(m1)
confint(m1)
```

**Interpretation**:

-   **Fixed intercept** ($\mu_0$): 1.62891, Average antisocial behavior score at age 6 (centered age = 0).
-   **Fixed slope** ($\mu_1$): 0.07425, Average annual rate of change across childhood. A positive value indicates behavior increases over time on average.
-   **Random intercept variance** ($\tau_{00}$): 1.01338, variance among the intercepts, between-person differences in starting level, also corresponding to the **.sig01** (Random Intercept Std.Dev.)
-   **Random slope variance** ($\tau_{11}$): 0.02295, variance among the slopes, Children differ meaningfully in their rate of change — some increase faster, others plateau or decline, **.sig03** (Random Slope Std.Dev.)
-   **Intercept–slope covariance** ($\tau_{01}$): 0.48, positive though not significant (C.I. from -0.02 to 1.00), higher starters accelerate further. **.sig02 (Correlation)**

```{r}
#| label: growth-plot
ggplot(antisocial_long, aes(age, predict(m1))) +
  geom_line(aes(group = id), alpha = 0.25, linewidth = 0.3, color = "grey40") +
  stat_summary(fun = mean, geom = "line",
               linewidth = 1.4, color = "steelblue") +
  labs(
    x     = "Centered Age (0 = age 6)",
    y     = "Predicted Antisocial Behavior",
    title = "Individual Growth Trajectories (grey) with Mean Trajectory (blue)"
  ) +
  theme_minimal()
```



------------------------------------------------------------------------

## Step 3 — Moderation by Sex

**Research question**: Do boys and girls differ in their baseline level of antisocial behavior *and* in their rate of change over time?

We add `male` as a level-2 predictor of both the random intercept and random slope, yielding a cross-level interaction: `age × male`.

$$y_{it} = (\mu_0 + \beta_0\,\text{male}_i) + (\mu_1 + \beta_1\,\text{male}_i)\cdot\text{age}_{it} + b_{0i} + b_{1i}\cdot\text{age}_{it} + \varepsilon_{it}$$

```{r}
#| label: sex-moderation
m2 <- lmer(
  anti ~ age * male + (1 + age | id),
  data    = antisocial_long,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
summary(m2)
```

### Simple Slopes: Sex-Specific Growth Rates

```{r}
#| label: sex-slopes
sim_slopes(
  model      = m2,
  pred       = age,
  modx       = male,
  centered   = "none",
  cond.int   = TRUE,
  johnson_neyman = FALSE
)
```

### Interaction Plot

```{r}
#| label: sex-interaction-plot
interact_plot(
  model       = m2,
  pred        = age,
  modx        = male,
  x.label     = "Centered Age (0 = age 6)",
  y.label     = "Predicted Antisocial Behavior",
  legend.main = "Sex",
  modx.labels = c("Girls", "Boys")
) +
  theme_minimal()
```

### Individual Trajectory Plots by Sex

```{r}
#| label: sex-trajectory-plots
#| fig-width: 10
#| fig-height: 4

antisocial_long$pred_m2 <- predict(m2)

p_girls <- ggplot(
  filter(antisocial_long, male == 0),
  aes(x = age, y = anti)
) +
  ylim(0, 10) +
  geom_line(aes(y = pred_m2, group = id),
            alpha = 0.3, linewidth = 0.3, color = "tomato") +
  stat_summary(aes(y = pred_m2), fun = mean, geom = "line",
               linewidth = 1.2, color = "darkred") +
  labs(title = "Girls", x = "Centered Age", y = "Antisocial") +
  theme_minimal()

p_boys <- ggplot(
  filter(antisocial_long, male == 1),
  aes(x = age, y = anti)
) +
  ylim(0, 10) +
  geom_line(aes(y = pred_m2, group = id),
            alpha = 0.3, linewidth = 0.3, color = "steelblue") +
  stat_summary(aes(y = pred_m2), fun = mean, geom = "line",
               linewidth = 1.2, color = "navy") +
  labs(title = "Boys", x = "Centered Age", y = "Antisocial") +
  theme_minimal()

gridExtra::grid.arrange(p_girls, p_boys, ncol = 2)
```

**Interpretation**:

-   $\beta_0$ (male on intercept): If positive and significant, boys start with higher antisocial behavior at age 6 than girls.
-   $\beta_1$ (age × male): If positive and significant, boys increase at a faster rate than girls over time. Simple slopes decompose this interaction into sex-specific slopes with standard errors.
-   **Practical significance**: Even if both trajectories are increasing, the *difference in slopes* indicates differential developmental risk.

------------------------------------------------------------------------

## Step 4 — Moderation by Home Cognitive Environment

**Research question**: Does the home cognitive environment (a continuous, standardized score) predict children's initial level and rate of growth in antisocial behavior?

```{r}
#| label: homecog-moderation
m3 <- lmer(
  anti ~ age * homecog + (1 + age | id),
  data    = antisocial_long,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
summary(m3)
```

### Simple Slopes at Three Levels of Home Cognitive Support

```{r}
#| label: homecog-slopes
# −1 SD, Mean, +1 SD of homecog
sim_slopes(
  model        = m3,
  pred         = age,
  modx         = homecog,
  modx.values  = c(-1, 0, 1),
  cond.int     = TRUE,
  johnson_neyman = FALSE
)
```

```{r}
#| label: homecog-interaction-plot
interact_plot(
  model        = m3,
  pred         = age,
  modx         = homecog,
  modx.values  = c(-1, 0, 1),
  modx.labels  = c("−1 SD", "Mean", "+1 SD"),
  x.label      = "Centered Age (0 = age 6)",
  y.label      = "Predicted Antisocial Behavior",
  legend.main  = "Home Cognitive\nEnvironment"
) +
  theme_minimal()
```

**Interpretation**:

-   $\beta_0$ (homecog on intercept): If negative and significant, children from more cognitively stimulating homes start lower on antisocial behavior.
-   $\beta_1$ (age × homecog): If negative and significant, a richer home cognitive environment buffers against the age-related *increase* in antisocial behavior — children from enriched environments show attenuated growth.
-   **Simple slopes** at −1 SD, mean, +1 SD allow readers to see the growth trajectory separately for low, average, and high home cognitive support.

------------------------------------------------------------------------

## Step 5 — Multiple-Group Model (Sex as Grouping Factor)

The cross-level interaction approach in Model 2 assumes homogeneous within-group residual variance and random-effect variances across boys and girls. The multiple-group model relaxes this, allowing separate $\tau_{00}$, $\tau_{11}$, $\tau_{01}$, and $\sigma^2$ for each sex.

```{r}
#| label: multigroup-prep
antisocial_long$female    <- ifelse(antisocial_long$male == 0, 1, 0)
antisocial_long$femaleXage <- antisocial_long$female * antisocial_long$age
antisocial_long$maleXage   <- antisocial_long$male   * antisocial_long$age

# Constrained model (equal residual variance across sex) — baseline for LRT
m2b <- lme(anti ~ age * male,
           random = ~ 1 + age | id,
           data   = antisocial_long)

# Multiple-group model: separate RE covariance matrices + residual variance
m4 <- lme(
  anti ~ age * male,
  random  = list(id = pdBlocked(list(
    ~ -1 + male    + maleXage,
    ~ -1 + female  + femaleXage
  ))),
  weights = varIdent(form = ~ 1 | male),
  data    = antisocial_long
)
summary(m4)
```

```{r}
#| label: residual-variances
cat("Residual variance (boys):  ", round(m4$sigma^2, 4), "\n")
sigma_female <- m4$sigma * exp(as.numeric(m4$modelStruct$varStruct))
cat("Residual variance (girls): ", round(sigma_female^2, 4), "\n")
```

### Likelihood Ratio Test: Multiple-Group vs. Constrained

```{r}
#| label: lrt
anova(m2b, m4)
```

**Interpretation**: If the LRT is significant ($p < .05$), the data provide evidence that boys and girls differ not only in their mean trajectories (captured by the interaction terms) but also in the *variability* of their trajectories — i.e., the between-person spread around the sex-specific mean growth curve differs by sex.

------------------------------------------------------------------------

## Model Comparison Summary

| Model | Fixed Effects | Random Effects | $\sigma^2$ |
|----------------|-------------------|----------------------|----------------|
| M0: Null | Intercept only | Random intercept | Homogeneous |
| M1: Unconditional growth | Intercept + Age | Intercept + Slope | Homogeneous |
| M2: Sex moderation | \+ Male + Age×Male | Intercept + Slope | Homogeneous |
| M3: HomeCog moderation | \+ HomeCog + Age×HomeCog | Intercept + Slope | Homogeneous |
| M4: Multiple-group | Intercept + Age×Male | Sex-specific | By sex |

------------------------------------------------------------------------

------------------------------------------------------------------------

# Part II: LSEM — Longitudinal Measurement Invariance Testing

## Background

Before interpreting whether a psychological construct *changes* over time, we must first verify that the construct is measured equivalently at each time point. This is the question of **longitudinal measurement invariance**: does the scale measure the same thing, in the same units, with the same origin, at each wave?

**Dataset**: `selfconcept.dat` — 228 children, self-concept measured at ages 8, 10, and 12 **Indicators** (4 per wave):

-   `flgd` — feeling good about oneself
-   `worth` — sense of personal worth
-   `satis` — satisfaction with oneself
-   `nogd` — not feeling no good (reversed)

**Model sequence** (nested, increasingly constrained):

| Step | Model           | Added Constraint            | Comparison        |
|------|-----------------|-----------------------------|-------------------|
| 1    | Configural      | Loadings free across time   | — (baseline)      |
| 2    | Weak (Metric)   | Equal loadings across time  | vs. Configural    |
| 3    | Strong (Scalar) | Equal loadings + intercepts | vs. Weak          |
| 4    | Partial Scalar  | Free misfit intercepts      | vs. Strong & Weak |

## Setup and Data Preparation

```{r}
#| label: lsem-setup
library(lavaan)

# Variable names (4 indicators × 3 time points)
sc_names <- c("flgd08","worth08","satis08","nogd08",
              "flgd10","worth10","satis10","nogd10",
              "flgd12","worth12","satis12","nogd12")

# Published means and SDs
sc_mns <- c(3.20, 3.30, 3.21, 2.67,
            3.21, 3.28, 3.12, 2.71,
            3.24, 3.32, 3.16, 2.88)

sc_sds <- c(0.57, 0.63, 0.64, 0.88,
            0.59, 0.60, 0.66, 0.87,
            0.61, 0.70, 0.68, 0.76)

# Read correlation matrix and convert to covariance matrix
sc_cor <- read.table("data/selfconcept.dat",
                     header    = FALSE,
                     row.names = sc_names,
                     col.names = sc_names)
sc_cor <- data.matrix(sc_cor)
sc_cov <- cor2cov(sc_cor, sds = sc_sds)
```

## Step 1 — Configural Invariance (Baseline)

The configural model establishes that the same factor structure holds at each time point: each wave has one latent self-concept factor indicated by the same four items. No cross-time equality constraints are imposed on loadings or intercepts — only the pattern of which items load on which factor is constrained.

Unique features of longitudinal CFA:

-   **Correlated residuals** across time for the same item (e.g., `flgd08 ~~ flgd10`) to account for item-specific stability beyond the latent factor.
-   **Latent factor means** at Time 1 fixed to 0 (identification); Times 2 and 3 estimated.
-   **Latent factor variances** at Time 1 fixed to 1 (scale-setting); Times 2 and 3 estimated.

```{r}
#| label: configural
config_mod <- '
  # --- Factor structure ---
  sc08 =~ NA*flgd08 + L.f*flgd08 + worth08 + satis08 + nogd08
  sc10 =~ NA*flgd10 + L.f*flgd10 + worth10 + satis10 + nogd10
  sc12 =~ NA*flgd12 + L.f*flgd12 + worth12 + satis12 + nogd12

  # --- Latent means (Time 1 = 0 for identification) ---
  sc08 ~ 0*1
  sc10 ~ start(0)*1
  sc12 ~ start(0)*1
  flgd08 ~ I.f*1;  flgd10 ~ I.f*1;  flgd12 ~ I.f*1

  # --- Latent variances (Time 1 = 1 for scale-setting) ---
  sc08 ~~ 1*sc08
  sc10 ~~ start(1)*sc10
  sc12 ~~ start(1)*sc12

  # --- Correlated residuals (same item across time) ---
  flgd08 ~~ flgd10 + flgd12;  flgd10 ~~ flgd12
  worth08 ~~ worth10 + worth12; worth10 ~~ worth12
  satis08 ~~ satis10 + satis12; satis10 ~~ satis12
  nogd08  ~~ nogd10  + nogd12;  nogd10  ~~ nogd12
'

fit_config <- sem(config_mod,
                  sample.cov  = sc_cov,
                  sample.mean = sc_mns,
                  sample.nobs = 228,
                  meanstructure = TRUE)

summary(fit_config, fit.measures = TRUE, estimates = TRUE)
```

**Interpretation**: The configural model serves as the *unconstrained baseline*. If fit is acceptable (CFI \> .95, RMSEA \< .06, SRMR \< .08), we proceed to add constraints. Note the pattern of loadings: all four items should load substantially (\> .40) on their respective time-point factors.

------------------------------------------------------------------------

## Step 2 — Weak (Metric) Invariance: Equal Factor Loadings

Weak invariance constrains all four factor loadings to be equal across the three time points (e.g., the loading of `worth` on self-concept is the same at ages 8, 10, and 12). This ensures that the *unit of measurement* for the latent factor is consistent over time — a prerequisite for comparing relationships (e.g., correlations, regressions) involving the construct across waves.

```{r}
#| label: weak
weak_mod <- '
  # Loadings constrained equal across time via shared labels
  sc08 =~ NA*flgd08 + L.f*flgd08 + L.w*worth08 + L.s*satis08 + L.n*nogd08
  sc10 =~ NA*flgd10 + L.f*flgd10 + L.w*worth10 + L.s*satis10 + L.n*nogd10
  sc12 =~ NA*flgd12 + L.f*flgd12 + L.w*worth12 + L.s*satis12 + L.n*nogd12

  sc08 ~ 0*1;  sc10 ~ start(0)*1;  sc12 ~ start(0)*1
  flgd08 ~ I.f*1;  flgd10 ~ I.f*1;  flgd12 ~ I.f*1

  sc08 ~~ 1*sc08;  sc10 ~~ start(1)*sc10;  sc12 ~~ start(1)*sc12

  flgd08 ~~ flgd10 + flgd12;  flgd10 ~~ flgd12
  worth08 ~~ worth10 + worth12; worth10 ~~ worth12
  satis08 ~~ satis10 + satis12; satis10 ~~ satis12
  nogd08  ~~ nogd10  + nogd12;  nogd10  ~~ nogd12
'

fit_weak <- sem(weak_mod,
                sample.cov  = sc_cov,
                sample.mean = sc_mns,
                sample.nobs = 228,
                meanstructure = TRUE)

summary(fit_weak, fit.measures = TRUE)
```

### Likelihood Ratio Test: Weak vs. Configural

```{r}
#| label: lrt-weak-config
lavTestLRT(fit_weak, fit_config)
```

**Interpretation**: A non-significant $\chi^2$ difference ($p > .05$) and negligible $\Delta$CFI (\< .01) indicate that constraining loadings to equality does not significantly worsen model fit — **weak invariance holds**. This means the scale's metric (unit of measurement) is stable over time.

------------------------------------------------------------------------

## Step 3 — Strong (Scalar) Invariance: Equal Loadings + Intercepts

Strong invariance adds the constraint that *item intercepts* are also equal across time points. This ensures that individuals with the same latent self-concept score will, on average, endorse each item to the same degree regardless of measurement occasion — a necessary condition for meaningfully comparing **latent means** across waves.

```{r}
#| label: strong
strong_mod <- '
  sc08 =~ NA*flgd08 + L.f*flgd08 + L.w*worth08 + L.s*satis08 + L.n*nogd08
  sc10 =~ NA*flgd10 + L.f*flgd10 + L.w*worth10 + L.s*satis10 + L.n*nogd10
  sc12 =~ NA*flgd12 + L.f*flgd12 + L.w*worth12 + L.s*satis12 + L.n*nogd12

  sc08 ~ 0*1;  sc10 ~ start(0)*1;  sc12 ~ start(0)*1
  flgd08 ~ I.f*1;  flgd10 ~ I.f*1;  flgd12 ~ I.f*1
  worth08 ~ I.w*1; worth10 ~ I.w*1; worth12 ~ I.w*1
  satis08 ~ I.s*1; satis10 ~ I.s*1; satis12 ~ I.s*1
  nogd08  ~ I.n*1; nogd10  ~ I.n*1; nogd12  ~ I.n*1

  sc08 ~~ 1*sc08;  sc10 ~~ start(1)*sc10;  sc12 ~~ start(1)*sc12

  flgd08 ~~ flgd10 + flgd12;  flgd10 ~~ flgd12
  worth08 ~~ worth10 + worth12; worth10 ~~ worth12
  satis08 ~~ satis10 + satis12; satis10 ~~ satis12
  nogd08  ~~ nogd10  + nogd12;  nogd10  ~~ nogd12
'

fit_strong <- sem(strong_mod,
                  sample.cov  = sc_cov,
                  sample.mean = sc_mns,
                  sample.nobs = 228,
                  meanstructure = TRUE)

summary(fit_strong, fit.measures = TRUE)
```

### LRT: Strong vs. Weak

```{r}
#| label: lrt-strong-weak
lavTestLRT(fit_strong, fit_weak)
```

### Modification Indices for Intercept Constraints

```{r}
#| label: mod-indices
mi <- lavTestScore(fit_strong)
mi$uni[order(-mi$uni$X2), ][1:10, ]
```

**Interpretation**: If the LRT is significant, scalar invariance does not fully hold. Modification indices highlight which specific intercept constraints are most problematic. Large M.I.s for a particular item's intercept at one time point suggest that item has a different response tendency at that occasion, independent of the latent construct level.

------------------------------------------------------------------------

## Step 4 — Partial Scalar Invariance

When full scalar invariance fails, we can achieve **partial scalar invariance** by freeing only the intercept constraints flagged by modification indices, while retaining equality constraints on the remaining items. Partial invariance is acceptable for latent mean comparisons *as long as at least two items per factor retain equal intercepts*.

Based on modification indices, the intercepts for `satis08` (age 8) and `nogd12` (age 12) are freed.

```{r}
#| label: partial-scalar
partial_mod <- '
  sc08 =~ NA*flgd08 + L.f*flgd08 + L.w*worth08 + L.s*satis08 + L.n*nogd08
  sc10 =~ NA*flgd10 + L.f*flgd10 + L.w*worth10 + L.s*satis10 + L.n*nogd10
  sc12 =~ NA*flgd12 + L.f*flgd12 + L.w*worth12 + L.s*satis12 + L.n*nogd12

  sc08 ~ 0*1;  sc10 ~ start(0)*1;  sc12 ~ start(0)*1
  flgd08 ~ I.f*1;  flgd10 ~ I.f*1;  flgd12 ~ I.f*1
  worth08 ~ I.w*1; worth10 ~ I.w*1; worth12 ~ I.w*1
  satis08 ~ 1        # freed: satis08 intercept allowed to differ
  satis10 ~ I.s*1;  satis12 ~ I.s*1
  nogd08  ~ I.n*1;  nogd10  ~ I.n*1
  nogd12  ~ 1        # freed: nogd12 intercept allowed to differ

  sc08 ~~ 1*sc08;  sc10 ~~ start(1)*sc10;  sc12 ~~ start(1)*sc12

  flgd08 ~~ flgd10 + flgd12;  flgd10 ~~ flgd12
  worth08 ~~ worth10 + worth12; worth10 ~~ worth12
  satis08 ~~ satis10 + satis12; satis10 ~~ satis12
  nogd08  ~~ nogd10  + nogd12;  nogd10  ~~ nogd12
'

fit_partial <- sem(partial_mod,
                   sample.cov  = sc_cov,
                   sample.mean = sc_mns,
                   sample.nobs = 228,
                   meanstructure = TRUE)

summary(fit_partial, fit.measures = TRUE, estimates = TRUE,
        standardized = TRUE, rsquare = TRUE)
```

### LRT Comparisons for Partial Model

```{r}
#| label: lrt-partial
# Partial vs. Strong: did freeing two intercepts significantly improve fit?
lavTestLRT(fit_strong, fit_partial)

# Partial vs. Weak: is partial scalar still more constrained than weak?
lavTestLRT(fit_partial, fit_weak)
```

**Interpretation of the two tests**:

-   **Partial vs. Strong**: Should be significant — freeing the two intercepts meaningfully improves fit.
-   **Partial vs. Weak**: Should be non-significant — the partial model fits equivalently to weak invariance, confirming that the freed intercepts account for the misfit in the strong model.

------------------------------------------------------------------------

## Model Comparison Summary

```{r}
#| label: fit-table
fit_stats <- function(fit, label) {
  fi <- fitMeasures(fit, c("chisq","df","pvalue","cfi","rmsea","srmr"))
  data.frame(Model = label,
             χ2    = round(fi["chisq"], 2),
             df    = fi["df"],
             p     = round(fi["pvalue"], 3),
             CFI   = round(fi["cfi"],   3),
             RMSEA = round(fi["rmsea"], 3),
             SRMR  = round(fi["srmr"],  3))
}

do.call(rbind, list(
  fit_stats(fit_config,  "1. Configural"),
  fit_stats(fit_weak,    "2. Weak (Metric)"),
  fit_stats(fit_strong,  "3. Strong (Scalar)"),
  fit_stats(fit_partial, "4. Partial Scalar")
))
```

**Reading the table**:

-   **CFI \> .95** and **RMSEA \< .06** indicate acceptable absolute fit at each step.
-   Moving from Configural → Weak: $\Delta$CFI \< .01 and non-significant $\Delta\chi^2$ → **weak invariance supported**.
-   Moving from Weak → Strong: significant $\Delta\chi^2$ → **full scalar invariance rejected**.
-   Moving from Strong → Partial: significant $\Delta\chi^2$ improvement → **partial scalar invariance accepted**.
-   **Partial scalar** fit comparable to **Weak**: confirms two freed intercepts were the only source of non-invariance.

**Substantive conclusion**: Self-concept as measured by this four-item scale is **partially scalar invariant** across ages 8–12. Three of four item intercepts are stable over time, satisfying the minimum requirement for latent mean comparisons. The freed items (`satis08` and `nogd12`) may reflect genuine age-related shifts in how children interpret these specific items — a substantively interesting finding in its own right.

------------------------------------------------------------------------

# Session Information

```{r}
#| label: session-info
sessionInfo()
```
